---
title: Forecasting Target Corporation's Quarterly Revenue
author: Zachary Dyne
date: '2019-10-31'
slug: target-quarterly-revenue-forecasting
categories:
  - MBA
tags:
  - Forecasting
  - Finance
subtitle: ''
summary: 'This post goes through a few forecasting methods to analyze and predict quarterly revenue for the Target Corporation. This post is a version of a project I worked on during my MBA program with Roxanne Adams and Matt Stone.'
authors: []
lastmod: '2020-05-28T15:10:22-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: yes
    df_print: kable
    toc_depth: 2
---

During my MBA program I had the opportunity to take a course on busines and economic forecasting. The revenue cycle of many businesses can be highly seasonal, which adds a degree of complexity to creating forecasts. This is especially true in the retail sector, which made forecasting Target Corporation's quarterly revenue from 1993-2001 a fun challenge! 

```{r message=F, warning=F}
#setup
library(tidyverse)
library(fpp2)
library(kableExtra)
library(ggthemes)
TargetLabF <- read_csv("C:/Users/Lenovo/Desktop/AGSM/Fall2019/Forecasting/TargetDecomp/TargetLabF.csv")
TargetLabF$Period <- with(TargetLabF, paste(Yr, Quarter, sep = ":Q"))

# Create Timeseries Data for Decomposition Analysis
TargetTS <- TargetLabF %>%
  filter(!is.na(Sales)) %>%
  select(Time, Sales) %>%
  ts(., start = c(1993, 1), end = c(2001, 4), frequency = 4)

# Plot Data
TargetLabF %>%
  filter(!is.na(Sales)) %>%
  ggplot(aes(x=Time, y=Sales)) + 
  geom_line(size = 2) + 
  scale_x_continuous(breaks = TargetLabF$Time, labels = TargetLabF$Period) +
  theme_economist_white() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Target Quarterly Sales 1993:Q1 - 2001:Q4") +
  ylab("Sales (in millions of US Dollars)")
```

Target's quarterly sales show a general upward trend with strong seasonality. There is a strong increase in Quarter 4 in each year, which is typical for the retail industry given the surge in holiday shopping. There isn't a clear cycle, though, meaning that the trend is fairly constnat. 

Since the seasonal component here is constant from year to year, a classical decomposition method should work nicely here. We will work first through an additive decomposition and then through a multiplicative decomposition process. 

## Additive Decomposition
```{r}
TargetAddDecompm <- TargetTS %>% 
  decompose(type = "additive")

TargetAddDecompm %>%
  autoplot() +
  ggtitle("Classical Additive Decomposition of Target Sales") +
  xlab("Year") +
  theme_economist_white() # choose economist theme
```
This plot is reasonably interesting, as it shows th trend and seasonally adjusted data alonogside each other, along with the error from the decomposition model. 


```{r}
# Extract Seasonal and Trend Components
Seasonal <- TargetAddDecompm$seasonal  # Extract seasonal indecies            
SeasAdj <- (TargetTS/Seasonal)         # Seasonally adjusted data   
TimeF <- TargetLabF %>% select(Time)   # Extract TimeF
TrendMod <- lm(SeasAdj ~ TargetTS)     # Fit trend to seasonally adjusted data   
TimeFramef <- TargetLabF %>% select(Sales) # Extract sales
Trend <- predict(TrendMod,              # Project trend using trend model  
                 newdata = TargetTS)
TrendDecomp <- TargetAddDecompm$trend   #Extract trend value
TrendDecompDF <- TrendDecomp[,2] %>% as.data.frame() # Convert trend value to df
names(TrendDecompDF) <- c("Trend") # rename

# to find model coeficients
summary(TrendMod)
```
 

```{r message = F, warning=F}
TrendVals <- Trend %>% 
  as.data.frame() %>% 
  select(TargetTS.Sales) %>% 
  ts(., start = c(1993, 1), end = c(2001, 4), frequency = 4)

#create aggregate dataframe for plotting and performance anlaysis
Forecast <- TrendVals * Seasonal
SalesData <- TargetTS %>% as.data.frame() %>% select(Sales) 
TimeData <- TargetTS %>% as.data.frame() %>%  select(Time)
AddPlot <- data.frame(TrendDecompDF, Forecast, SalesData, TimeData)
names(AddPlot) <- c("Trend", "Forecast", "Data", "Time")
AddPlot %>%
  gather(key="DataType", value="Value", 1:3) %>%
  ggplot(aes(x=Time, y=Value, color = DataType)) + 
  geom_line(size=1.5) + 
  theme_economist_white() +
  scale_x_continuous(breaks = TargetLabF$Time, labels = TargetLabF$Period) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Additve Decomposition", subtitle = "Forecast and Actual Data") +
  ylab("Sales (in millions of US Dollars)")
```

Seasonal Indecies (in millions of dollars)
```{r warning=F}
Seasonal %>%
  as.tibble() %>%
  rowid_to_column() %>%
  filter(rowid < 5) %>%
  rename("Quarter" = "rowid") %>%
  rename("Seasonal_Indicator" = "x")
```

Performance Metrics
```{r}
# create all of the relevant performance metrics
SSTO <- sum((AddPlot$Data - mean(AddPlot$Data))^2)
SSE <- sum((AddPlot$Data - AddPlot$Forecast)^2)
MSD <- SSE/length(AddPlot$Data)
s <- sqrt(MSD)
R2 <- (SSTO - SSE)/SSTO
MAD <- sum(abs(AddPlot$Data - AddPlot$Forecast))/length(AddPlot$Data)
MPE <- 100*sum((AddPlot$Data - AddPlot$Forecast)/AddPlot$Data)/length(AddPlot$Data)
MAPE <- 100*sum(abs((AddPlot$Data - AddPlot$Forecast)/AddPlot$Data))/length(AddPlot$Data)  
Performance <- tibble(SSTO, SSE, MSD, s, R2, MAD, MPE, MAPE)
```

#### Additive Decomposition Performance Metrics
```{r}
Performance %>% kable() %>% kable_styling()
```

### Additive Decomposition Assessment

Through all of that, our predicted spike in quarter four is expected to continue, but the ability of the forecast to predict actual data in quater 3 in years 1996-1998 are far off, as the plot of forecast would suggest.  

We have a farily low R2, which is the portion of variation explained by the model. This coincides with a high MAD (Mean Absolute Deviation) and SSE (Sum of Squered Error), which are both metrics of by how much the model misses actual data. While these are useful quantitative metircs, the of forecast against actual data indicates that perhaps we should try multiplicative decomposition methods for forecasting Target's quarterly revenue. 

## Multiplicative Decomposition
```{r}
TargetMultDecomp <- TargetTS %>% 
  decompose(type = "multiplicative") 

TargetMultDecomp %>%
  autoplot() +
  ggtitle("Classical Multiplicative Decomposition of Target Sales") +
  xlab("Year") + 
  theme_economist_white()
```

```{r warning=F}
# Extract Seasonal and Trend Components
Seasonal <- TargetMultDecomp$seasonal              
SeasAdj <- (TargetTS/Seasonal)          
TimeF <- TargetLabF %>% select(Time)
TrendMod <- lm(SeasAdj ~ TargetTS)
TimeFramef <- TargetLabF %>% select(Sales)
Trend <- predict(TrendMod,                
                 newdata = TargetTS)
TrendDecomp <- TargetMultDecomp$trend
TrendDecompDF <- TrendDecomp[,2] %>% as.data.frame()
names(TrendDecompDF) <- c("Trend")

TrendVals <- Trend %>% 
  as.data.frame() %>% 
  select(TargetTS.Sales) %>% 
  ts(., start = c(1993, 1), end = c(2001, 4), frequency = 4)

#create aggregate dataframe for plotting and performance anlaysis
Forecast <- TrendVals * Seasonal
SalesData <- TargetTS %>% as.data.frame() %>% select(Sales) 
TimeData <- TargetTS %>% as.data.frame() %>%  select(Time)
MultPlot <- data.frame(TrendDecompDF, Forecast, SalesData, TimeData)
names(MultPlot) <- c("Trend", "Forecast", "Data", "Time")
MultPlot %>%
  gather(key="DataType", value="Value", 1:3) %>%
  ggplot(aes(x=Time, y=Value, color = DataType)) + 
  geom_line(size=1.5) + 
  scale_x_continuous(breaks = TargetLabF$Time, labels = TargetLabF$Period) +
  theme_economist_white() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Multiplicative Decomposition") +
  ylab("Sales (in millions of US Dollars)")
```

```{r}
Seasonal %>%
  as.tibble() %>%
  rowid_to_column() %>%
  filter(rowid < 5) %>%
  rename("Quarter" = "rowid") %>%
  rename("Seasonal_Indicator" = "x")
```

```{r}
SSTO <- sum((MultPlot$Data - mean(MultPlot$Data))^2)
SSE <- sum((MultPlot$Data - MultPlot$Forecast)^2)
MSD <- SSE/length(MultPlot$Data)
s <- sqrt(MSD)
R2 <- (SSTO - SSE)/SSTO
MAD <- sum(abs(MultPlot$Data - MultPlot$Forecast))/length(MultPlot$Data)
MPE <- 100*sum((MultPlot$Data - MultPlot$Forecast)/MultPlot$Data)/length(MultPlot$Data)
MAPE <- 100*sum(abs((MultPlot$Data - MultPlot$Forecast)/MultPlot$Data))/length(MultPlot$Data)  
Performance <- tibble(SSTO, SSE, MSD, s, R2, MAD, MPE, MAPE)
```

#### Multiplicative Decomposition Performance Metrics
```{r}
Performance %>% kable() %>% kable_styling()
```

### Multiplicative Decomposition Assessment
The multiplicative model performs noteably better than the additive model. The R2 indicates that almost 99% of the varaition is captured by the model. As demonstrated by the low MAD and SSE values, the model has a fairly low total error. This model may actually be useful for forecasting Target's 2002 and 2003 quarterly revenue. Let's do it! 

## Forecasting 2002 - 2003
```{r}
TargetTS %>% 
  forecast() %>%
  autoplot() +
  ggtitle("Target Sales Forecast Using Multiplicative Decomposition") + 
  ylab("Sales & Forecast (in millions of US dollars") +
  theme_economist_white()
```

```{r}
TargetTS %>% 
  forecast() %>%
  as.data.frame() %>% 
  filter(Series == "Sales") %>%
  select(-Series)
```


